Where Do Events Come From?

### 2018-08-01 paper.tex
...

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
    code, public, private,
    var, watching, when, with},
}
\lstset{
  literate={~} {$\sim$}{1},
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\footnotesize\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\tiny,
  %xleftmargin=0.5cm,
  %xrightmargin=0.5cm,
  %upquote=true,
}

@@

%% Title information
\title{Where Do Events Come From?}
\subtitle{Reactive Programming From The Ground Up}

@@

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} indicates this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In the example, even if two packets arrive simultaneously in the environment,
the synchronous model provides the atomicity and responsiveness guarantees:
    the first \code{await} will awake and turn the LED off atomically
    (ln 12--13);
    and the second \code{await} will awake in sequence (ln 14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

\section{Interrupt Service Routines in \CEU}
\label{sec.isrs}

@@

Asynchronous execution confronts the synchronous mindset of \CEU since the
assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Nevertheless, our goal is to preserve the well-behaved interaction between the
\CEU application and the environment as of Figure~\ref{fig.env}, even in the
presence of asynchronous ISRs as illustrated in Figure~\ref{fig.async}:
%
the original application is now represented by the synchronous \CEU and the
environment by the asynchronous \CEU, all inside the same application.
%
This way, we want to push all subtleties of asynchronous execution into device
drivers which emit input events to regular synchronous code in \CEU exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;
input  high/low PIN_02;
#include "gpio.ceu"

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.gpio.async},
  escapechar=\%,
]
// OUTPUT DRIVER

@@

input high/low PIN_02;
{
    pinMode(2, INPUT_PULLUP);
    EICRA = (EICRA & ~((1<<ISC00) | (1<<ISC01)))
          | (CHANGE << ISC00);
    EIMSK |= (1<<INT0);
}
spawn async/isr [INT0_vect] do
    emit PIN_02({digitalRead(2)});
end
\end{lstlisting}

Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} implements an application
that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
is on whenever the button is pressed.
%
The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
the external world and includes the asynchronous side as a driver (ln 1--3).
%
The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
input events.

An \code{output} implementation (ln 4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln 3,5) with interpolation to
evaluates \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries in embedded
toolchains, such as Arduino.
%
In the example, the driver sets pin 13 as output (ln 3) when it is included,
and sets its state whenever it is emitted (ln 5).

An \code{input} event implementation requires an ISR registered with the
\code{spawn async/isr} primitive (ln 17--19), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will always have an \code{emit} to an input event inside its
body to awake the synchronous side.
However, although the ISR executes asynchronously as soon as the interrupt
occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
an ongoing reaction on the synchronous side.
%
In the example, the input driver first configures pin 2 to behave as input and
to generate an external interrupt on level transitions (ln 11--16).

synchronous remains simple and well behaved
with no low-level calls

The application only communicates

>>>>

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
    code, public, private,
    var, watching, when, with},
}
\lstset{
  literate={~} {$\sim$}{1},
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\footnotesize\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  numbersep=4pt,
  numberstyle=\tiny,
  %xleftmargin=0.5cm,
  %xrightmargin=0.5cm,
  %upquote=true,
}

@@

%% Title information
\title{Where Do Events Come From?}
\subtitle{Reactive and Energy-Efficient Programming From The Ground Up}

@@

The synchronous execution model of \CEU dictates that reactions to input events
are atomic and that incoming events are never lost.
%
The event loop in Figure~\ref{fig.env} indicates this behavior since the
environment can only generate a new input after the application yields control
and closes the loop.
%
In the example, even if two packets arrive simultaneously in the environment,
the synchronous model provides the atomicity and responsiveness guarantees:
    the first \code{await} will awake and turn the LED off atomically
    (ln~12--13);
    and the second \code{await} will awake in sequence (ln~14).
%
%Note that if multiple trails in parallel await the same input, they will all
%awake from it during the same reaction.
%In this case, \CEU dictates that each trail executes atomically, one after the
%other, in lexical order.
%This rule prevents race conditions for trails in parallel.

@@

\subsection{Interrupt Service Routines}
\label{sec.isrs}

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Nevertheless, our goal is to preserve the well-behaved interaction between the
\CEU application and the environment as of Figure~\ref{fig.env}, even in the
presence of asynchronous ISRs as illustrated in Figure~\ref{fig.async}:
%
the original application is now represented by the synchronous \CEU and the
environment by the asynchronous \CEU, all inside the same application.
%
This way, we want to push all subtleties of asynchronous execution into device
drivers which emit input events to regular synchronous code in \CEU exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;     // connected to an LED
input  high/low PIN_02;     // connected to a button
#include "gpio.ceu"

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Asynchronous driver for GPIO which, on hardware interrupts, emits an
           input event into a queue.},
  label={lst.gpio.async},
  escapechar=\%,
]
// OUTPUT DRIVER

@@

input high/low PIN_02;
{
    pinMode(2, INPUT_PULLUP);
    EICRA |= (1 << ISC00);  // sets INT0
    EIMSK |= (1 << INT0);   // turns on INT0
}
spawn async/isr [INT0_vect] do
    emit PIN_02({digitalRead(2)});
end
\end{lstlisting}

Listing~\ref{lst.gpio.sync}~and~\ref{lst.gpio.async} implements an application
that uses GPIO to link a button (pin 02) to an LED (pin 13) such that the LED
is on whenever the button is pressed and off whenever it is unpressed.

The synchronous side (Listing~\ref{lst.gpio.sync}) declares its interface with
the external world and includes the asynchronous side as a driver (ln~1--3).
It then starts with the LED off (ln~5) and enters in a loop (ln~6--9) that,
whenever the button changes (ln~7), changes the state of the LED with the new
value (ln~8).

The asynchronous side (Listing~\ref{lst.gpio.async}) implements the output and
input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluates \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries in embedded
toolchains, such as Arduino.
%
In the example, the driver sets pin 13 as output (ln~3) when it is included,
and sets its state whenever it is emitted (ln~5).
%
An \code{input} event implementation requires an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically have an \code{emit} to an input event inside its
body to awake the synchronous side.
However, although the ISR executes asynchronously as soon as the interrupt
occurs, the \code{emit} respects the semantics of \CEU and does not interrupt
an ongoing reaction on the synchronous side (going into a queue to be
discussed).
%
In the example, the input driver first configures pin 2 to behave as input and
to generate an external interrupt on level transitions (ln~11--15).

The example illustrates the clear separation between the application and the
driver through an event-driven interface that reflects the desired architecture
of Figures~\ref{fig.env}~and~\ref{fig.async}.
%
Now the whole application is written in \CEU:
    the synchronous side remains well behaved with no low-level calls, and
    the asynchronous side deals with all complexity in device drivers.
%
Nevertheless, drivers are typically write-once components developed by embedded
specialists, which are reused in most applications.
%
Note that each architecture requires a mapping between the \code{async/isr} and
the actual interrupt vector table, which we provide with our system.

As we discuss in the next section, supporting ISRs in the same language and
memory space of the application allows programmers to choose between efficiency
and safety during development by providing (and experimenting with) multiple
interfaces with different levels of abstraction.

\subsection{Safe Shared-Memory Concurrency}
\label{sec.ceu.shared}

\begin{figure}[t!]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[
  numbers=right,
  xrightmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.a},
  caption={Shared \code{x}},
  escapechar=\%,
]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[
  xleftmargin=3em,
  basicstyle=\ttfamily\small,
  label={lst.mem.b},
  caption={Shared \code{y}},
  escapechar=\%,
]
input void A;
// (empty line)
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

\end{lstlisting}
\hskip1.5em
\begin{minipage}[t]{.9\columnwidth}
\end{minipage}%
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{
Accesses to shared \code{x} never concurrent.
Accesses to shared \code{y} are concurrent but still deterministic.},
\label{lst.mem}
\end{figure}

In \CEU, when multiple trails awake in the same reaction, they execute in
lexical order, i.e., in the order they appear in the program source code.
%
In Figure~\ref{lst.mem}, both examples define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously in accordance with the synchronous model.
Hence, for the sequence of events \code{A->B}, \code{x} becomes
\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order, the execution is still
deterministic, and \code{y} always becomes \code{4} (\code{(1+1)*2}).

Regardless of deterministic execution, note that an apparently innocuous change
in the order of trails may modify the behavior of programs.
%
To mitigate this threat, \CEU performs concurrency checks at compile time to
detect conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
The static checks are optional (or raise warnings) and do not affect the
semantics of the language.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART buffer is shared between the ISR and the application,
           resulting in a pontential race condition.
           \emph{(The driver and application code are in separate files.)}},
  label={lst.usart},
]
// USART INTERFACE

input none USART_RX;
var[32*] byte rx_buf;             // `*' = circular

// DRIVER

spawn async/isr [USART_RX_vect] do
    rx_buf = rx_buf .. [{UDR0}];
    emit USART_RX;
end

// APPLICATION

loop do
    await USART_RX;
    var int i;
    loop i in [0 -> $rx_buf[ do   // `$' = length ($)
        // uses rx_buf[i]
    end
    $rx_buf = 0;
end
\end{lstlisting}

The addition of asynchronous ISRs poses real threats concerning race conditions
in programs since interrupts occur at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a circular buffer to avoid missing data (ln~4).
The buffer is indispensable because the microcontroller cannot cope with the
USART speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signal the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

\CEU treats an \code{async/isr} as a block of code that runs in parallel with
the rest of the program (hence the required prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts (supposedly) for a short period
of time.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART driver now exposes a safer (and more friendly) interface
           to the application.
           \emph{(The driver and application code are in separate files.)}},
  label={lst.usart.code},
]
// USART INTERFACE

code Usart_RX (var&[] byte buf, var int n) -> none;

// DRIVER

input none USART_RX;
var[32*] byte rx_buf;

spawn async/isr [USART_RX_vect] do
    rx_buf = rx_buf .. [{UDR0}];
    emit USART_RX;
end

code Usart_RX (var&[] byte buf, var int n) -> none
do
    loop do
        atomic do
            buf = buf..rx_buf;
            $rx_buf = 0;
        end
        if $buf >= n then
            break;
        end
        await USART_RX;
    end
end

// APPLICATION

loop do
    var[255] byte buf;
    await Usart_RX(&buf, 10);
    // uses buf
end
\end{lstlisting}

Even so, we do not expect that application programmers are required to resolve
race conditions.
\CEU supports reactive abstractions (similar to coroutines) that help hiding
the concurrency complexity inside the driver.
%
Listing~\ref{lst.usart.code} changes the USART API (ln 1--3) to expose a code
abstraction that expects a reference to a buffer and a number of bytes to
receive.
Now, the application (ln~29--35) invokes the abstraction (ln~33) with a local
buffer (ln~32).
%
The driver (ln~5--27) now hides the low level interface of the previous example
(ln~7--8) and implements the code abstraction (ln~15--27) that protects the
access to the shared buffer with an \code{atomic} block (ln~18--21).
%
The abstraction copies the driver buffer into the application buffer (ln~19)
until the receiving the request number of bytes (ln~22).
%
On the one hand the extra copying incurs an overhead, but on the other hand it
prevents race conditions with the ISR.

\subsection{Energy-Efficient Runtime}
\label{sec.energy}

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.},
  label={lst.env},
]
evt_t queue[MAX];           // input queue
void main () {
  ceu_start();              // "boot reaction"
  while (1) {
    evt_t evt;
    if (ceu_input(&evt)) {  // query input queue
      ceu_sync(&evt);       // execute synchronous
    } else {
      ceu_pm_sleep();       // nothing to execute
    }
  }
}
\end{lstlisting}

Our runtime of \CEU implements the event loop proposed in
Figure~\ref{fig.async}, which alternates between synchronous, well-behaved
execution, and asynchronous ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, we want the runtime to enter in sleep mode automatically to
save energy.

Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program to spawn the ISRs and reach the first \code{await}
statements in the multiple lines of code.
%
Then, the runtime enters an infinite loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop, but they typically only
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

Microcontrollers typically support multiple levels of sleep mode, each
progressively saves more energy at the expense of keeping less functionality
active.
%
As an example, the least efficient mode of the \emph{ATmega328P} allows for
timer interrupts since it keeps its internal clock active, while the most
efficient mode turns off all peripherals and can only awake from external
interrupts.

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying the vector every time
\code{ceu\_pm\_sleep} is called.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={XXX.
           \emph{(The power manager, driver and application code are in separate files.)}},
  label={lst.pm},
  escapechar=\%,
]
// POWER MANAGER (in C)

enum {
    CEU_PM_ADC,
    CEU_PM_TIMER1,
    CEU_PM_USART,
    ...
};

void ceu_pm_sleep (void) {
    if (ceu_pm_get(CEU_PM_USART) || ...) {
        sleep_mode_1(...);
    } else if (ceu_pm_get(CEU_PM_ADC)) {
        sleep_mode_2(...);
    } else {
        sleep_mode_3(...);
    }
}

// USART DRIVER (in Ceu)

code Usart_RX (var&[] byte buf, var int n) -> none
do
    {ceu_pm_set(CEU_PM_USART, 1);}
    do finalize with
        {ceu_pm_set(CEU_PM_USART, 0);}
    end
    ...   // same as in Listing 7
end

// APPLICATION (in Ceu)

input high/low PIN_02;      // connected to a button
loop do
    par/or do
        await PIN_02;
    with
        var[255] byte buf;
        await Usart_RX(&buf, 10);
        // uses buf
    end
    await PIN_02;
end
\end{lstlisting}

The third configuration achieves optimal efficiency but requires a tight
interaction between the drivers and the power management runtime.
%
Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
driver (ln~20--29), and an illustrative application (ln~31--43).
%
The application is a loop that awaits in parallel for a button press (ln~36) or
receiving 10 bytes (ln~39).
Let's call this initial state \code{STATE-A}.
The \code{par/or} terminates when either of them occurs, going to the next line
that awaits another button click (ln~42).
Let's call this other state \code{STATE-B}.
After the button click, the program loops back to \code{STATE-A}.
%
While in \code{STATE-A}, the program can awake from USART and external
interrupts, which means that the power manager should choose a sleep mode in
which the USART remains operational.
%
While in \code{STATE-B}, only external interrupts should awake the program,
which menas that the power manager may use the most efficient sleep mode.
%
The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
before sleeping, queries their states (ln~11,13) to choose the most appropriate
sleep mode (ln~12,14,16).
%
The USART driver needs to be extended with calls to enable and disable the
USART state inside the power manager: just before awaiting, the driver enables
the USART (ln~24) and registers a finalization block to unregister it on
termination (ln~25--27).
Termination may occur either directly after receiving the requested number of
bytes, or indirectly if the \code{par/or} in the application terminates on a
button click (ln~36).

Note that applications need not to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous-reactive execution model
and energy-aware runtime of \CEU.

\section{Related Work}

- TinyOS
    - async, race condition avoidance
        - nesc manual
    - ICEM, standby
        - icem paper

- CRP

- finalize
- structured programming
    - lexical scoping vs manual
    - no turn off

- citar ICEM o mais cedo nas duas ocasioes (async, energy)
- citar TOS/ICEM na introducao como maior influencia
    - ja no abstract?

- falar que async/isr nao contem nada de Ceu, apenas shared memory e nocao de estar em paralelo
    - compartilhamento de memoria
        - velocidade
    - ajuda do compilador p/ detectar acesso concorrente
